---
title: "Getting Started with fastmlm"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with fastmlm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

**fastmlm** fits linear mixed-effects models using the same formula syntax as
`lme4::lmer()`, but with a high-performance C++ backend that delivers 4-14x
speedups. It provides:

- A custom formula parser (3.8x faster than `lme4::lFormula`)
- Formula/data caching for repeated fits (bootstrap, simulation)
- C++ L-BFGS-B optimizer with forward-difference gradients
- CHOLMOD supernodal sparse Cholesky (via R's Matrix package)
- Precomputed sparse matrix mappings (zero-allocation deviance updates)
- Optional GPU acceleration via CUDA (cuBLAS/cuSPARSE)
- Satterthwaite degrees of freedom (no lmerTest dependency needed)
- Compatibility with emmeans and broom.mixed

## Basic usage

fastmlm uses the same formula interface as lme4:

```{r basic}
library(fastmlm)
library(lme4)
data(sleepstudy)

m <- fmlm(Reaction ~ Days + (Days | Subject), data = sleepstudy)
m
```

## Extracting results

All the standard methods work:

```{r extractors}
fixef(m)
head(ranef(m)$Subject)
VarCorr(m)
```

Variance-covariance of fixed effects:

```{r vcov}
vcov(m)
```

Confidence intervals:

```{r confint}
confint(m)
```

## Comparing with lme4

fastmlm produces numerically identical results:

```{r compare}
m_lme4 <- lmer(Reaction ~ Days + (Days | Subject), data = sleepstudy)

# Fixed effects agree to machine precision
max(abs(fixef(m) - lme4::fixef(m_lme4)))

# Residual SD
c(fastmlm = sigma(m), lme4 = sigma(m_lme4))
```

## REML vs ML

```{r reml-ml}
m_reml <- fmlm(Reaction ~ Days + (1 | Subject), data = sleepstudy, REML = TRUE)
m_ml   <- fmlm(Reaction ~ Days + (1 | Subject), data = sleepstudy, REML = FALSE)

c(REML_sigma = sigma(m_reml), ML_sigma = sigma(m_ml))
c(REML_AIC = AIC(m_reml), ML_AIC = AIC(m_ml))
```

## Crossed random effects

fastmlm handles crossed designs (common in psycholinguistics):

```{r crossed}
set.seed(42)
n_subj <- 30; n_item <- 20
subj <- factor(rep(1:n_subj, each = n_item))
item <- factor(rep(1:n_item, times = n_subj))
x <- rnorm(n_subj * n_item)
y <- 2 + 0.5 * x + rnorm(n_subj, sd = 1.5)[subj] +
     rnorm(n_item, sd = 1.0)[item] + rnorm(n_subj * n_item)
df <- data.frame(y, x, subj, item)

m_crossed <- fmlm(y ~ x + (1 | subj) + (1 | item), data = df)
m_crossed
```

## Formula caching for repeated fits

When fitting the same model repeatedly (e.g., bootstrapping), fastmlm
caches the parsed formula and design matrices. The second fit onwards
skips all formula parsing:

```{r caching}
# First fit (cold)
system.time(fmlm(Reaction ~ Days + (Days | Subject), data = sleepstudy))

# Second fit (cached — formula parsing skipped)
system.time(fmlm(Reaction ~ Days + (Days | Subject), data = sleepstudy))

# Clear cache if needed
fastmlm_clear_cache()
```

## System information

Check what backends are active:

```{r sysinfo}
fastmlm_blas_info()
```

## Control parameters

Fine-tune the optimizer:

```{r control}
m <- fmlm(Reaction ~ Days + (Days | Subject), data = sleepstudy,
          control = fmlm_control(
            optimizer = "L-BFGS-B",
            maxiter = 500,
            ftol = 1e-12,
            gtol = 1e-10
          ))
```

## Working with emmeans

fastmlm integrates with emmeans for estimated marginal means:

```{r emmeans, eval=requireNamespace("emmeans", quietly=TRUE)}
library(emmeans)
em <- emmeans(m, ~ Days, at = list(Days = c(0, 5, 9)))
em
pairs(em)
```

## Nonlinear effects with restricted cubic splines

fastmlm supports `rms::rcs()`, `splines::ns()`, and other basis functions
directly in the formula. The spline basis is constructed by the external
package; fastmlm fits the (still linear-in-parameters) mixed model:

```{r splines, eval=requireNamespace("rms", quietly=TRUE)}
library(rms)
set.seed(42)
n <- 300; subj <- factor(rep(1:30, each = 10))
age <- runif(n, 20, 80)
y <- 5 + 0.01 * (age - 50)^2 + rnorm(30, sd = 3)[subj] + rnorm(n, sd = 2)
d <- data.frame(y = y, age = age, subj = subj)

m_rcs <- fmlm(y ~ rcs(age, 4) + (1 | subj), data = d)
m_rcs

# Test nonlinearity via likelihood ratio test
m_lin <- fmlm(y ~ age + (1 | subj), data = d, REML = FALSE)
m_nl  <- fmlm(y ~ rcs(age, 4) + (1 | subj), data = d, REML = FALSE)
anova(m_lin, m_nl)
```

Prediction with new data works correctly --- knot positions from the
training data are automatically reused:

```{r spline-predict, eval=requireNamespace("rms", quietly=TRUE)}
predict(m_rcs, newdata = data.frame(age = c(30, 50, 70)))
```

## Working with broom.mixed

Tidy output is available:

```{r broom}
fastmlm:::tidy.fmlmMod(m)
fastmlm:::glance.fmlmMod(m)
```
