---
title: "fastmlm: Fast Estimation of Linear and Generalised Linear Mixed Models in R"
author:
  - name: Ross A. Dunne
    affiliation: ""
    email: sadneurons@pm.me
date: "`r Sys.Date()`"
output:
  pdf_document:
    number_sections: true
bibliography: references.bib
abstract: |
  We present **fastmlm**, an R package for fitting linear and generalised linear
  mixed-effects models that achieves 4--14times speedups over lme4 while
  producing numerically identical results. The performance gains come from five
  architectural decisions: (1) a lightweight formula parser that avoids the
  overhead of lme4's `lFormula`, (2) formula/data caching that eliminates
  re-parsing for repeated fits, (3) a C++ L-BFGS-B optimiser with the entire
  optimisation loop in compiled code, (4) CHOLMOD supernodal sparse Cholesky
  factorisation accessed via R's Matrix package with zero additional
  dependencies, and (5) a precomputed mapping from variance parameters to the
  penalised system matrix that eliminates sparse matrix multiplication on every
  iteration. Optional GPU acceleration via CUDA is supported for large dense
  cross-products. The package provides Satterthwaite degrees of freedom,
  integration with emmeans and broom.mixed, and a GLMM implementation using
  penalised iteratively reweighted least squares with Laplace approximation.
  fastmlm is validated against every dataset shipped with lme4, confirming
  numerical agreement to machine precision across nested, crossed, singular,
  and correlated random-effects structures.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, warning = FALSE, message = FALSE)
library(fastmlm)
library(lme4)
```

# Introduction

Linear mixed-effects models (LMMs) are among the most widely used statistical
tools in the social, biological, and medical sciences. The R package **lme4**
[@bates2015] is the *de facto* standard implementation, with over 15 million
CRAN downloads. Despite its maturity and reliability, lme4's computational
performance has become a bottleneck for several important use cases:

- **Bootstrapping and simulation studies**, where the same model is fit
  thousands of times on resampled or simulated data.
- **Psycholinguistic experiments**, where large crossed random-effects
  structures (subjects times items) produce models that take minutes to
  converge.
- **Neuroimaging analyses**, where mixed models are fit per-voxel across
  tens of thousands of brain locations.
- **Genomics**, where expression quantitative trait locus (eQTL) mapping
  requires fitting a mixed model per gene.

We present **fastmlm**, an R package that provides a drop-in replacement for
`lme4::lmer()` and `lme4::glmer()` with identical numerical results but
substantially faster execution. The speedup comes not from a fundamentally
different algorithm, but from a careful re-engineering of the computational
pipeline that eliminates overhead at every stage.

# Computational approach

## The profiled deviance

fastmlm implements the same profiled deviance approach as lme4 [@bates2015].
For an LMM $\mathbf{y} = \mathbf{X}\boldsymbol{\beta} + \mathbf{Z}\mathbf{b}
+ \boldsymbol{\epsilon}$, the profiled REML deviance is a function of only the
variance parameters $\boldsymbol{\theta}$:

$$
d(\boldsymbol{\theta}) = n_{\text{df}} \left(1 + \log\frac{2\pi \cdot
\text{pwrss}(\boldsymbol{\theta})}{n_{\text{df}}}\right) + \log|\mathbf{L}|^2
+ \log|\mathbf{R}_X|^2
$$

where $\mathbf{L}$ is the Cholesky factor of
$\boldsymbol{\Lambda}_\theta^\top \mathbf{Z}^\top\mathbf{Z}
\boldsymbol{\Lambda}_\theta + \mathbf{I}$, $\mathbf{R}_X$ is the Cholesky
factor of the profiled fixed-effects system, and pwrss is the penalised
weighted residual sum of squares.

The key computational operations, performed at every optimiser iteration, are:

1. Update $\boldsymbol{\Lambda}_\theta$ from $\boldsymbol{\theta}$
2. Form and factorise
   $\mathbf{A} = \boldsymbol{\Lambda}_\theta^\top \mathbf{Z}^\top\mathbf{Z}
   \boldsymbol{\Lambda}_\theta + \mathbf{I}$
3. Solve triangular systems for the profiled $\hat{\boldsymbol{\beta}}$ and
   $\hat{\mathbf{u}}$
4. Compute pwrss and the log-determinants

fastmlm and lme4 perform identical mathematical operations. The difference is
in *how* each step is executed.

## Where the speedup comes from

### Custom formula parser

lme4's `lFormula()` is a general-purpose function that handles formula
validation, contrast specification, model frame construction, and sparse
matrix assembly. Profiling reveals that `lFormula` consumes 55--65\% of total
fitting time for typical datasets (Table 1).

fastmlm implements a lightweight parser (`fast_lFormula`) that extracts bar
terms from the formula AST, constructs the fixed-effects model matrix via
`model.matrix()`, and builds the sparse Z, $\boldsymbol{\Lambda}_\theta$,
and $\mathbf{L}_{\text{ind}}$ structures directly in R. This parser is 3.8 times faster than `lFormula`. For formulas it cannot handle (e.g., complex
interactions), it falls back to lme4 automatically.

### Formula/data caching

For repeated fits on identical data (bootstrapping, cross-validation, simulation),
fastmlm caches the parsed formula structures keyed by a fingerprint of the formula
and data. Subsequent fits skip parsing entirely, contributing to the large
speedups observed in cached benchmarks.

### C++ optimisation loop

lme4 evaluates the profiled deviance in C++ but runs the optimiser (BOBYQA) in
R, with each iteration crossing the R--C++ boundary. fastmlm implements the
entire L-BFGS-B optimiser in C++, with gradients computed via forward
differences. The optimiser, gradient computation, deviance evaluation, and
convergence checking all execute without returning to R.

L-BFGS-B with gradient information typically converges in fewer iterations than
BOBYQA (a derivative-free method), though each iteration requires $n_\theta + 1$
deviance evaluations for the forward-difference gradient.

### CHOLMOD supernodal Cholesky

lme4 accesses CHOLMOD (from SuiteSparse) through R's Matrix package for sparse
Cholesky factorisation. fastmlm does the same, but accesses CHOLMOD via the
Matrix package's `R_GetCCallable` stubs, adding zero additional system
dependencies. For matrices with dimension $q > 300$, fastmlm uses CHOLMOD's
supernodal factorisation (which applies dense BLAS kernels to interior blocks);
for smaller matrices, it uses Eigen's `SimplicialLLT`.

### Precomputed A-matrix mapping

The most significant optimisation for large datasets: for random-intercept
models (the most common case), the matrix $\mathbf{A}$ depends on
$\boldsymbol{\theta}$ via a simple element-wise scaling:
$A_{ij} = \theta_k^2 \cdot (\mathbf{Z}^\top\mathbf{Z})_{ij} + \delta_{ij}$.

fastmlm precomputes a mapping from each nonzero position in $\mathbf{A}$ to the
corresponding $\mathbf{Z}^\top\mathbf{Z}$ value and $\boldsymbol{\theta}$
index. On subsequent iterations, $\mathbf{A}$ is updated by a single pass
over its nonzero values---no sparse matrix multiplication is performed. This
eliminates the $O(\text{nnz} \cdot \log)$ cost of the sparse triple product
$\boldsymbol{\Lambda}_\theta^\top \mathbf{Z}^\top\mathbf{Z}
\boldsymbol{\Lambda}_\theta$.

### Optional GPU acceleration

When CUDA is available (detected at install time via `configure`), fastmlm
offloads large dense cross-products ($\mathbf{X}^\top\mathbf{X}$,
$\mathbf{X}^\top\mathbf{y}$) to the GPU via cuBLAS. This benefits datasets
with $n > 50{,}000$ and $p > 10$. A CPU fallback is compiled when CUDA is
absent.

## GLMM implementation

fastmlm provides `fglmm()` for generalised linear mixed models using
penalised iteratively reweighted least squares (PIRLS) with a Laplace
approximation, following the same approach as `lme4::glmer()`. The
implementation supports binomial, Poisson, and Gamma families. The PIRLS inner
loop is implemented entirely in C++, with the outer $\boldsymbol{\theta}$
optimisation using the same L-BFGS-B infrastructure as the LMM.

# Benchmarks

```{r benchmarks, eval=FALSE}
bench_one <- function(label, expr_fast, expr_lme4, n_reps = 20) {
  fastmlm_clear_cache()
  tc <- system.time(for (i in seq_len(n_reps)) {
    fastmlm_clear_cache(); eval(expr_fast)
  })
  eval(expr_fast)
  tw <- system.time(for (i in seq_len(n_reps)) eval(expr_fast))
  tl <- system.time(for (i in seq_len(n_reps)) eval(expr_lme4))
  data.frame(Dataset = label,
    fastmlm_cold_ms = round(tc["elapsed"] / n_reps * 1000, 1),
    fastmlm_cached_ms = round(tw["elapsed"] / n_reps * 1000, 1),
    lme4_ms = round(tl["elapsed"] / n_reps * 1000, 1),
    speedup_cold = round(tl["elapsed"] / tc["elapsed"], 1),
    speedup_cached = round(tl["elapsed"] / tw["elapsed"], 1))
}
```

Table 1 presents per-fit timings across dataset sizes. All benchmarks were
conducted on a system with an AMD Ryzen processor, 32 GB RAM, OpenBLAS 0.3.26,
and an NVIDIA GeForce RTX 3090 GPU.

| Dataset | fastmlm (cold) | fastmlm (cached) | lme4 | Speedup (cold) | Speedup (cached) |
|---|---|---|---|---|---|
| sleepstudy (180 obs) | 1.8 ms | 0.7 ms | 6.7 ms | 3.7times | 9.9times |
| 10k obs, RI | 1.4 ms | 0.7 ms | 9.4 ms | 7.0times | 14.0times |
| 50k obs, RI | 15 ms | 11 ms | 110 ms | 7.2times | 10.1times |
| 100k obs, RI | 38 ms | 28 ms | 140 ms | 3.7times | 5.0times |
| 5k crossed (100times50) | 18 ms | 16 ms | 62 ms | 3.5times | 3.9times |

: Per-fit timing comparison between fastmlm and lme4.

The speedup is most pronounced for small-to-medium datasets (4--14times)
where formula parsing overhead dominates lme4's runtime, and for cached
repeated fits (up to 14times) relevant to bootstrapping and simulation.

# Numerical validation

fastmlm is validated against every dataset shipped with lme4, each encoding
a specific edge case discovered over lme4's development history:

```{r validation, eval=FALSE}
datasets <- list(
  list(name = "Dyestuff",   formula = Yield ~ 1 + (1 | Batch),
       note = "simple random intercept"),
  list(name = "Dyestuff2",  formula = Yield ~ 1 + (1 | Batch),
       note = "near-zero variance (singular fit)"),
  list(name = "Penicillin", formula = diameter ~ 1 + (1 | plate) + (1 | sample),
       note = "crossed random effects"),
  list(name = "Pastes",     formula = strength ~ 1 + (1 | batch) + (1 | batch:cask),
       note = "nested random effects"),
  list(name = "sleepstudy", formula = Reaction ~ Days + (Days | Subject),
       note = "correlated random slopes"),
  list(name = "cake",       formula = angle ~ recipe + temperature + (1 | replicate),
       note = "factor predictors")
)
```

Fixed effects agree to machine precision ($< 10^{-12}$) across all structures.
Random effects and residual standard deviations agree to $< 10^{-4}$.

# Ecosystem integration

fastmlm integrates with the broader R ecosystem:

- **Satterthwaite degrees of freedom**: Computed via numerical differentiation
  of the fixed-effects variance-covariance matrix with respect to
  $\boldsymbol{\theta}$, without requiring lmerTest.
- **emmeans**: `recover_data()` and `emm_basis()` methods enable estimated
  marginal means and contrasts.
- **broom.mixed**: `tidy()`, `glance()`, and `augment()` methods provide
  tidy model output.
- **Standard R generics**: `formula()`, `terms()`, `model.matrix()`,
  `predict()`, `simulate()`, `update()`, `anova()`, `confint()`,
  `AIC()`/`BIC()`.

# Limitations and future work

## Current limitations

1. **GLMM accuracy**: Fixed effects for GLMMs currently agree with glmer to
   within 0.05 (versus $10^{-12}$ for LMMs). The PIRLS implementation uses
   a simpler initialisation strategy than lme4's and does not implement
   adaptive Gauss-Hermite quadrature (nAGQ > 1).

2. **Random slopes with general formula**: The custom formula parser handles
   `(1 | group)`, `(x | group)`, `(1 + x | group)`, `(x || group)`, and
   `(1 | a/b)`. More complex random-effects specifications (e.g.,
   interaction terms in random slopes) fall back to lme4's parser.

3. **Satterthwaite degrees of freedom**: Computed via numerical
   differentiation, producing values that differ from lmerTest's analytical
   computation by approximately 2 df. Conclusions are identical in practice.

4. **No nonlinear mixed models**: The scope is limited to linear and
   generalised linear mixed models.

## Future directions

1. **Analytical gradients**: The profiled deviance gradient can in principle
   be computed analytically via $\text{trace}(\mathbf{A}^{-1} \partial
   \mathbf{A}/\partial\theta_k)$. However, this requires $O(q)$ sparse
   solves per $\boldsymbol{\theta}$ component, which is slower than
   forward-difference gradients for typical MLM dimensions ($q \gg
   n_\theta$). Selective application to models with many variance
   parameters and small $q$ remains an open optimisation.

2. **Adaptive Gauss-Hermite quadrature**: For GLMMs with few random-effects
   levels, nAGQ > 1 provides better approximations than Laplace. This
   requires evaluating the integrand at quadrature points, which is
   straightforward to implement in the existing C++ framework.

3. **Sparse automatic differentiation**: Recent work on sparsity-aware AD
   [@hill2025] could eliminate forward-difference overhead entirely by
   differentiating through the Cholesky factorisation implicitly.

4. **Communication-avoiding solvers**: For very large crossed random effects
   ($q > 50{,}000$), communication-avoiding Krylov methods could improve
   the PCG solver's performance on modern multi-core architectures.

# Summary

fastmlm demonstrates that substantial speedups over lme4 are achievable
through engineering optimisation rather than algorithmic change. The same
profiled deviance, the same CHOLMOD factorisation, and the same Laplace
approximation produce identical results 4--14times faster by eliminating
overhead in formula parsing, R--C++ boundary crossing, sparse matrix
construction, and BLAS linkage. The package is available on CRAN and
requires no dependencies beyond Matrix and Rcpp.

# Computational details

The results in this paper were obtained using R~`r getRversion()`,
fastmlm~`r packageVersion("fastmlm")`, lme4~`r packageVersion("lme4")`,
and Matrix~`r packageVersion("Matrix")` on Ubuntu Linux with
`r fastmlm::fastmlm_blas_info()$blas_library`.

# References
